A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-06-14 bioc-workflow.Rmd
plotDist example

**TODO** Mention that there are other dimensionality reduction
techniques. Advantages, disadvantages? See
[this](https://gist.github.com/mikelove/74bbf5c41010ae1dc94281cface90d32)
very interesting gist with intriguing tSNE examples. This might be
better to come after markers.

@@

In the context of spatial proteomics, a marker protein is defined as a well-known resident of a specific sub-cellular niche in a species *and* condition of interest. Applying this to machine learning (ML), and specifically supervised learning, for the task of protein localisation prediction, markers constitute the labelled training data to use as input to a classification analyses. Defining well-known residents, and obtaining labelled training data for ML analyses can be time consuming, but it is important to define markers that are representative of the multivariate data space and on which a classifier will be trained and generated. `r Biocpkg("pRoloc")` provides a convenience function, `addMarkers`, to directly add markers to a `MSnSet` object, as demonstrated in the code chunk below. These marker sets can be accessed using the `pRolocmarkers()` function. Marker sets are stored as a simple named vector in R, and originate from in-house user-defined spreadsheets or a set of markers from previous published studies. The marker vectors that can be accessed from `pRolocmarkers` are named vectors and to enable mapping between the markers and the `MSnSet` instance it is required that the `featureNames` of the `MSnSet` instance match the `names` of the marker. The mouse dataset used here has Uniprot IDs stored as the `featureNames` (see `head(featureNames(lopit2016))`) and the names of the vector of the mouse markers (`mmus` markers) are also Uniprot IDs (see `head(mrk)` in the code chunk below), so it is straightforward to match names between the markers and the `MSnSet` instance. If the naming between the marker sets and the `MSnSet` dataset are different, one will have to convert and match the proteins according to the appropriate identifier. Sometimes, we find the equivalent entry name, Uniprot ID or accession number is stored with the data, which makes conversion between identifers relatively straightforward. If this is not the case however, there are conversion softwares online available, for example XXX.

In the code chunk below, we demonstrate how to add markers using `pRolocmarkers` function and then visualise these annotations using the `plot2D` function.

```{r addmrkers}
## List available marker sets
pRolocmarkers()

## Use mouse markers
mrk <- pRolocmarkers(species = "mmus")
head(mrk)

## Add mouse markers
hl <- addMarkers(hl, mrk)

## Plot mouse markers
plot2D(hl, fcol = "markers", main = "pRolocmarkers for mouse")
```

@@

@@

```{r}
i1 <- exprs(hl[, hl$replicate == 1])
i2 <- exprs(hl[, hl$replicate == 2])
pearcor <- cor(t(i1), t(i2))
mean(diag(pearcor))
```

>>>>

**TODO** Advantages, disadvantages? See
[this](https://gist.github.com/mikelove/74bbf5c41010ae1dc94281cface90d32)
very interesting gist with intriguing tSNE examples. This might be
better to come after markers.

@@

In the context of spatial proteomics, a marker protein is defined as a
well-known resident of a specific sub-cellular niche in a species
*and* condition of interest. Applying this to machine learning (ML),
and specifically supervised learning, for the task of protein
localisation prediction, markers constitute the labelled training data
to use as input to a classification analyses. Defining well-known
residents, and obtaining labelled training data for ML analyses can be
time consuming, but it is important to define markers that are
representative of the multivariate data space and on which a
classifier will be trained and generated. `r Biocpkg("pRoloc")`
provides a convenience function, `addMarkers`, to directly add markers
to a `MSnSet` object, as demonstrated in the code chunk below. These
marker sets can be accessed using the `pRolocmarkers()`
function. Marker sets are stored as a simple named vector in R, and
originate from in-house user-defined sets of markers or from previous
published studies [@Gatto:2014b].

```{r markers}
## List available marker sets
pRolocmarkers()
```

These markers can then be mapped to an `MSnSet`'s `featureNames`. The
mouse dataset used here has Uniprot IDs stored as the `featureNames`
(see `head(featureNames(lopit2016))`) and the names of the vector of
the mouse markers (`mmus` markers) are also Uniprot IDs (see
`head(mrk)` in the code chunk below), so it is straightforward to
match names between the markers and the `MSnSet` instance using the
`addMarkers` function.

```{r addmrkers}
## Use mouse markers
mrk <- pRolocmarkers(species = "mmus")
head(mrk)

## Add mouse markers
hl <- addMarkers(hl, mrk)
```

If the naming between the marker sets and the `MSnSet` dataset are
different, one will have to convert and match the proteins according
to the appropriate identifier. Sometimes, we find the equivalent entry
name, Uniprot ID or accession number is stored with the data, which
makes conversion between identifers relatively straightforward. If
this is not the case however, conversion can be performed using the 
`r Biocpkg("biomaRt")`, the Bioconductor
[annotation resouces](http://bioconductor.org/help/workflows/annotation/Annotation_Resources/)
or any conversion softwares available online.

We now visualise these annotations along the PCA plot using the
`plot2D` function. We also use the `addLegend` function to map the
marker classes to the pre-defined colours.

```{r plotmarkers}
## Plot mouse markers
plot2D(hl, fcol = "markers", main = "pRolocmarkers for mouse")
addLegend(hl, fcol = "markers", where = "bottomleft", cex = .7)
```

The colours have been defined so as to enable to differenciate up to
30 classes. If more are provided, different character symbols
(circles, squares, ... and empty and solid symbols) are used. The
colours and the default plotting characters (solid dots for the
markers and empty circles for the features of unknown localisation)
can of course be changed, as described in the `setStockcol` manual
page.

@@

Another useful visualisation that relies on marker annotation is the
representation of the protein profiles occupancy along the gradient
using the `plotDist` function. While the PCA plot enables to
efficiently visualise the complete dataset and assess the relative
separation of different sub-cellular niches, comparing profiles of a
few marker clusters is useful to assess how exactly they differ (in
terms of peak fractions, for example). Below, we plot the profile of
the mitochondrial and peroxisome markers.

```{r plotDist}
par(mfrow = c(2, 1))
o <- order(hyperLOPIT$Fraction.No)
cls <- c(Peroxisome = "darkblue", Mitochondrion = "purple")
for (k in c("Mitochondrion", "Peroxisome")) {
    plotDist(hyperLOPIT[fData(hyperLOPIT)$markers == k, o],
             pcol = cls[k])
    title(main = k)
}
```

@@

```{r}
i1 <- exprs(hl[, hl$replicate == 1])
i2 <- exprs(hl[, hl$replicate == 2])
pearcor <- cor(t(i1), t(i2))
pearcor0 <- cor(t(i1[, seq(1, 10, 2)]), t(i2[, seq(1, 10, 2)]))
mean(diag(pearcor))
mean(diag(pearcor0))
```

```{r}
data(hyperLOPIT2015ms3r1)
data(hyperLOPIT2015ms3r2)
data(hyperLOPIT2015ms3r3)

hll <- MSnSetList(list(hyperLOPIT2015ms3r1,
                       hyperLOPIT2015ms3r2,
                       hyperLOPIT2015ms3r3))

ij <- combn(3, 2)
pc <- apply(ij, 2,
            function(.ij) {
                tmp <- suppressMessages(commonFeatureNames(hll[.ij]))
                diag(cor(t(exprs(tmp[[1]])), t(exprs(tmp[[2]]))))
            })
names(pc) <- apply(ij, 2, paste, collapse = ".")
sapply(pc, summary)
```

The third replicate was produce over 1 year after the first two by
another operator.
